Transient Variability In Vapor Intrusion And The Factors That Influence It

### 2018-10-18 tex/main.tex
Updated some of Eric changes

\begin{abstract}
Temporal variability in indoor air contaminant concentrations at vapor intrusion (VI) sites has been a concern for some time.
We consider the source of the variability at VI sites located near Hill Air Force Base and Naval Air Station North Island using statistical analysis methods and three-dimensional subsurface computational fluid dynamics modeling.
The results suggest that an order of magnitude variation in indoor air contaminant concentrations may be expected at ”normal” VI sites where preferential pathways do not play a role, whereas three or more orders of magnitude can be observed at sites characterized by preferential pathways.
A CFD modeling sensitivity analysis reveals that it is not only the presence of contaminant vapor in a preferential pathway that is required to see large observed variations, but there must also exist a permeable region beneath the structure.
Large temporal fluctuations in indoor air contaminant concentrations may be observed where no preferential pathways exist, but this requires a particular combination of large fluctuations in pressure driving force and high soil permeability.
\end{abstract}

@@

Long term vapor intrusion (VI) studies in both residential and larger commercial structures have raised concerns regarding significant observed transient behavior in indoor air contaminant concentrations\cite{u.s._environmental_protection_agency_oswer_2015,folkes_observed_2009,holton_temporal_2013,johnston_spatiotemporal_2014,hosangadi_high-frequency_2017,mchugh_recent_2017}.
VI involves the migration of volatilizing contaminants from soil, groundwater or other subsurface sources into overlying structures. VI has been a recognized problem for some time, but many aspects remain poorly understood, particularly with respect to the causes of large temporal transients in indoor air concentrations.
There is uncertainty within the VI community regarding how to best develop sampling strategies to address this problem\cite{u.s._environmental_protection_agency_oswer_2015,holton_temporal_2013,johnson_integrated_2016}. \par

@@

Similarly, a sewer pipe has recently been suggested to be a source of the contaminants found in the EPA Indianapolis house\cite{mchugh_evidence_2017}.
That site was also characterized by large indoor air contaminant concentration fluctuations.
Sewer lines have been generally implicated as VI source at several sites to date\cite{pennell_sewer_2013,mchugh_evidence_2017,roghani_occurrence_2018,riis_vapor_2010}.
A Danish study estimates that roughly 20\% of all VI sites in central Denmark involve significant sewer VI pathways\cite{nielsen_remediation_2017}.
Thus while the consideration of a role of possible sewer or other preferential pathways is now part of normal good practice in VI site investigation, it is still not known whether the existence of such pathways automatically means that large temporal fluctuations are necessarily to be expected.
In some of these cases\cite{pennell_sewer_2013,riis_vapor_2010}, the sewer provided a pathway for direct entry of contaminant into the living space.
While potentially important in many cases, this scenario is not further considered here, where the focus is on pathways that deliver contaminant to the soil beneath a structure. \par

It is, however, now known that even absent a preferential pathway, there may be significant transient variation in indoor air contaminant concentrations at VI sites\cite{folkes_observed_2009,brenner_results_2010,johnston_spatiotemporal_2014}.
One example is a site at NAS North Island at which no preferential pathways have been identified.
Instead, a building at this site is characterized by significant variations in indoor-outdoor pressure differential\cite{hosangadi_high-frequency_2017}.
It is believed that this is the origin of the fluctuations at that site. \par

This paper investigates the sources of the temporal variation in indoor air contaminant concentrations in both the presence and absence of in soil preferential pathways.
In this work, the latter scenarios are referred to as ”normal” VI scenarios, in which there is typically a groundwater source of the contaminant.
Specifically, we pose the question of just how much variation in indoor air contaminant concentration may be expected at  such normal  VI  sites vs. those characterized by preferential pathways.
The conditions required for preferential pathways to be significant contributors to temporal variations in indoor air contaminant concentrations are also explored, and the consequences for sampling strategies are also discussed.

@@

The ASU house and data were obtained over a period of a few years.
During part of this time, controlled pressure method (CPM) tests were being conducted, in which the house was underpressurized to an extent greater than that characterizing “normal” operation.
This caused greater than normal advective flow from the subsurface into the house, thus increasing VI potential\cite{mchugh_evaluation_2012,mchugh_recent_2017,holton_evaluation_2015}.
This period of CPM testing is considered separately from the otherwise ”natural” VI conditions in the analysis.
Likewise, the existence of a preferential pathway at the ASU house needs to be considered in examining the dataset, noting that during some of the testing, this pathway was deliberately cut off, resulting in what we have termed “normal” VI conditions in which the main source of contaminant was believed to be groundwater. \par

@@

Likewise, the Indianapolis site investigation spanned over a number of years and periodically included the testing of a sub-slab depressurziation system (SSD).
The goal of which is to mitigate the VI risk by drastically depressurizing the sub-slab area underneath the house, preventing the contaminants from entering the structure above, and therefore only the period before the installation of this system was considered in the analysis.
It is likely a sewer line beneath the structure acted as a preferential pathway\cite{mchugh_evidence_2017}, however at no point was this PP removed, making it difficult to assess how significant the PP was at this site, regardless it is of interest to analyze this site due to its wealth of data. \par

The typical variation in indoor air contaminant concentrations with time will first be considered below in the case of the ASU house during ”natural”, (i.e. non-CPM conditions), in the case of the NAS North Island site over the entire available dataset, and for the Indianapolis case we consider the variations before the installation of the SSD system.
The deviations from the mean TCE (Chloroform and PCE at the Indianapolis site) in indoor air concentration, as well as the indoor-outdoor pressure differential associated with these concentration fluctuations, were examined, and both univariate and bivariate kernel density estimations (KDE) were constructed.
KDE is a technique that estimates the probability distribution of a random variable(s) by using multiple kernels, or weighting functions, and in this case, Gaussian kernels are used to create the KDEs.
This means that it is presumed that the variables of interest (i.e., indoor air contaminant concentrations and indoor-outdoor pressure differentials, as sampled) are normally distributed around mean values (that there is statistical fluctuation associated with each sampling event).
In this instance, the scipy statistical package was used to construct the KDEs, assuming a bandwidth parameter determined by Scott's rule.
The distributions of the individual parameters and the relationship between them will be examined using the KDE method.

\subsection{Modeling Work}
In addition to examining the actual field data, a previously described three -dimensional computational fluid dynamics model of a generic VI impacted house was used to elucidate certain aspects of  the processes.
This model was implemented in a finite element solver, COMSOL Multiphysics.
In the present work, there has been an addition of a preferential pathway to the ”standard” model that has been described before in publications by this group\cite{shen_influence_2013,yao_investigating_2017,yao_three-dimensional_2017}.
In this modeling work, only the vadose zone soil domain is directly modeled.

The modeled structure is assumed to have a 10x10 m foundation footprint, with the bottom of the foundation slab lying 1 m below ground surface (bgs), simulating a house with a basement.
The indoor air space is modeled as a continuously stirred tank (CST)\cite{u.s._environmental_protection_agency_oswer_2015} and all of the contaminant entering the house is assumed to enter with soil gas through a 1 cm wide crack located between the foundation walls and the foundation slab that spans the perimeter of the house.
All of the contaminant leaving the indoor air space is assumed does so via air exchange with the ambient.
The indoor control volume is assumed to consist of only of the basement, assumed as having a total volume of $300 \; \mathrm{m^3}$.
Clearly different assumptions could be made regarding the structural features and the size of the crack entry route, but for present purposes, this is unimportant as the intent is only to show for “typical” values what the influence of certain other features can be.
The modeled surrounding soil domain extends 5 meters from the perimeter of the house, and is assumed to consist of sandy clay (except as noted).
Directly beneath the foundation slab, there is assumed to be a 30 cm (one foot) thick gravel layer, except in certain cases where this sub-base material is assumed to be the same as the surrounding soil (termed  a ”uniform” soil scenario).
The preferential pathway is modeled as a 10 cm (4”) pipe that exits into the gravel sub-base beneath the structure.
The air in the pipe is assumed to be contaminated with TCE at a vapor concentration equal to the vapor in equilibrium with the groundwater contaminant concentration below the structure, modified by a scaling factor $\chi$, which is allows the contaminant concentration in the pipe to be parameterized.
The groundwater beneath the structure is assumed to be homogeneously contaminated with trichloroethylene (TCE) as a prototypical contaminant.
The groundwater itself is not modeled, as the bottom of the model domain is defined by the top of the water table.
The ground surface and the pipe are both assumed to be sources of air to the soil domain.
Both are assumed to be at reference atmospheric pressure.
Vapor transport in the soil is governed by Richard’s equation, a modified version  of Darcy’s Law, taking the variability of soil moisture in the vadose zone into account\cite{richards_capillary_1931}.
The van Genuchten equations are used to predict the soil moisture content and thus the effective permeability of the soil\cite{van_genuchten_closed-form_1980}.
The effective diffusivity of contaminant in soil is calculated using the Millington-Quirk model\cite{millington_permeability_1961}.
The transport of vapor contaminant in the soil is assumed to be governed by the advection-diffusion equation, in which either advection or diffusion may dominate depending upon position and particular circumstances.
The equations and the boundary conditions are given in Table \ref{tbl:eqns-bc-parameters}.

@@

\subsection{Drivers For Indoor Air Contaminant Variability}
\subsubsection{The Role Of Building Depressurization/Pressurization}
% KDE figures
\begin{figure}[htb!]
  \caption{KDE analysis of IACC dependence on indoor/outdoor pressure difference at the ASU house and North Island site (\ref{fig:kde-asu-nas}) and the Indianapolis site (\ref{fig:kde-indianapolis}). p-values and Pearson's r-values shown for each dataset.}
  \label{fig:kde-analysis}
  % ASU/North Island plot
  \begin{subfigure}{\textwidth}
    \caption{Period before and after the PP was shut at the ASU house considered separately; North Island dataset considered in its entirety.}
    \label{fig:kde-asu-nas}
    \includegraphics[height=0.4\textheight,keepaspectratio]{nas-asu-house-iacc-deviating.pdf}
  \end{subfigure}
  % Indianapolis plot
  \begin{subfigure}{\textwidth}
    \caption{Chloroform and PCE considered separeately at Indianapolis duplex 422.}
    \label{fig:kde-indianapolis}
    \includegraphics[height=0.4\textheight,keepaspectratio]{indianapolis-422.pdf}
  \end{subfigure}
\end{figure}
% Introduction
Out of the factors that influence IACC in VI, pressure is one of the most dynamic ones, and the relationship between changes in indoor/outdoor pressure difference and IACC are examined in Figure \ref{fig:kde-analysis}.
Three well-studied VI sites are considered, the "ASU house" near Hill AFB in Utah, a building at North Island NAS in California and a duplex in Indianapolis.\par
% Univariate IACC description
The absolute IACC at these sites vary significantly, therefore some means of comparing them all to each other is necessary.
Additionally, the focus in this section is the driver for variations in IACC, and representation of the variations between the different sites must also be achieved, specifically variations on the scale of order of magnitudes is of the greatest interest, as these are the largest hinderance to proper site IACC characterization.
To achieve these goals, the log-10 deviation from the mean IACC ($\mu$) within each dataset is calculated, e.g. $10\mu$ indicates the IACC is an order of magnitude above the mean IACC for the dataset.
On the y-margin in figures \ref{fig:kde-asu-nas} and \ref{fig:kde-indianapolis}, the univariate KDE distribution of deviation from the mean may be seen. \par
% Univariate pressure description
The indoor/outdoor pressure differences are simply taken as it, and their univariate KDE distributions are shown on the x-margin.
A negative pressure difference indicate that the building is underpressurized relative to outdoor.\par
% Bivariate description
The relationship between the deviation from mean IACC, and indoor/outdoor pressure difference may be seen in the central portion of each figure, in the form of a bivariate KDE distribution.
The p-value and Pearson's r-value for each bivariate distribution is shown in the legend.\par
% Sub-figure description (details about sites)
In Figure \ref{fig:kde-asu-nas} the North Island and ASU house datasets are plotted, with blue representing the North Island site, and with red and green representing the ASU house.
The ASU house dataset is split into two different parts, the first (red) is data from the time period before the PP was discovered, the second (green) is from after the PP was sealed off.
At both of these sites, TCE was the only contaminant considered.
The IACC of Chloroform and PCE in the 422 part of the Indianapolis duplex are considered in Figure \ref{fig:kde-indianapolis}.\par
% North Island discussion
From Figure \ref{fig:kde-asu-nas}, it is apparent that the three datasets differ significantly from each other.
The North Island site exhibit the greatest variation in both pressure difference and IACC, in fact, looking at the univariate pressure distribution - it is almost entirely flat and spanning from extremly high and low values.
A likely explation for this is the poor condition of the structure, rendering it highly susceptible to envivornmental influnces.\par

To accompany this, there is significant variation in IACC, that has a relativly strong link with pressure difference ($r = -0.71$), although there is clearly a distinct peak (around $0.5\mu$) that is primally associated with the smaller negative and positive pressure differences ($~-4 \leq \Delta p \leq 8$).
This demonstrates that at this particular site, the very large variations (an order of magnitude or more) are primarily influenced by the unusually large pressure difference fluctations.
Likely, this relationship is amplified by the fact that the soil surrounding the structure is sandy and therefore relatively permeable - increasing the influence of pressure differences.
However, despite all this, pressure differences can only partly explain the variations observed at this site.\par
% ASU discussion
The two ASU datasets are not only significantly different from the North Island dataset, but also different from each other.
Consider the dataset before the PP was discovered (red) and one can see that the IACC varies significantly (from $0.05 \mu$ to more than $50 \mu$, albeit rarely.)
This is more variation than was observed at North Island, and yet the distribution of pressure differences is narrow, mostly varying by a few Pa around zero (~$\pm 2$ Pa), and with $r = -0.50$, suggests that pressure difference is not insignificant but still a weak predictor for the variations in IACC.\par

The picture changes completely after the PP was closed off, where a similar distribution of pressure differences is observed, but with a signficant reduction in IACC variance to around $\pm 0.5 \mu$, resembling something of a log-normal distribution.
This clearly shows just how much variance the PP at this site contributed with, and how critical it is to assess whether a PP is present at a particular site in general, lest potentially face unacceptable uncertainty in determining the relevant contaminant exposure.
It is also clear that in the absence of a PP, the pressure difference becomes even less important, with the p- and r-value indicating that pressure differences are here insignficant in determining variation in IACC.
This also suggests that the PP at the ASU house acted not only as a PP for contaminant vapors, but for air in general - allowing the impermeable soil surrounding the house to be circumvented.
PPs, and the factors that influnce them will be explored later in this article.\par
% Indianapolis discussion
The Indianapolis duplex also featured a PP, but despite this, there is significantly less variance in IACC of both Chloroform and PCE.
The exact reason for this is not known, but considering that the PP at the Indianapolis site was a sanitary sewer, and at ASU the PP was a foundation drain system, it is not inconceivable that their dynamics are quite different.
Nevertheless, both PCE and Chloroform are more or less log-normally distributed, with (mostly) less than an order of magnitude variance around the mean, indicating that the varation in IACC are species independent.\par

The pressure difference distribution exhibit some bimodal characteristics, with a larger variation than was observed at the ASU site.
This may explain why the r-values for both PCE and Chloroform, $r=-0.51$ and $r=-0.53$ respectively, are so similar to the ASU site before the PP there was closed; the larger pressure differences, which occured more often than ASU, increase contaminant entry rates to the extent that a PP was needed to "achieve" at the ASU site.\par
% Summary discussion
All of these data clearly suggest that there is significant variation in IACC that cannot be easily explained by variations in indoor/outdoor pressure difference - even with a PP.
It is also clear that PP may contribute with unacceptable levels of variations, and screening for PPs should be part of any site investiation.
It may also be prudent, even in the absence of PP, to add around a half to one order of magnitude margin of error to early screening measurements of indoor air.
E.g. if a single sample is less than an order of magnitude from some target limit, it may be justified to perform subsequent samples to better assess the relevant exposure.
These data also call the role of advection in VI into question, as "normal" pressure difference ranges are such a poor predictor of IACC variation, which is the opposite of what one would expect if advection was a dominant transport mechanism.\par

@@

\begin{figure}[htb!]
  \caption{Sensitivity analysis of IACC dependence on indoor/outdoor pressure difference for a model featuring a PP (\ref{fig:ss-sensitivity-analysis-pp}) and without a PP (\ref{fig:ss-sensitivity-analysis-no-pp}). Results compared with field data from ASU house.}
  \label{fig:ss-sensitivity-analysis}
  % "Overview" plot
  \begin{subfigure}{0.55\textwidth}
    \caption{PP present. Sensitivity to the presence of a gravel sub-base and contamination in the PP considered.}
    \label{fig:ss-sensitivity-analysis-pp}
    \includegraphics[width=\textwidth]{yes-pp.pdf}
  \end{subfigure}
  % Violinplot figure
  \begin{subfigure}{0.55\textwidth}
    \caption{No PP preseent. Sensitivitiy to the presence of a gravel sub-base considered.}
    \label{fig:ss-sensitivity-analysis-no-pp}
    \includegraphics[width=\textwidth]{no-pp.pdf}
  \end{subfigure}
\end{figure}

>>>>

\begin{abstract}
Temporal variability in indoor air contaminant concentrations at vapor intrusion (VI) sites has been a concern for some time.
We consider the source of the variability at VI sites located near Hill Air Force Base in Utah, an EPA experimental house in Indianapolis, IN and Naval Air Station North Island in California using statistical analysis methods and three-dimensional subsurface computational fluid dynamics modeling.
The results suggest that an order of magnitude variation in indoor air contaminant concentrations may be expected at ”normal” VI sites where preferential pathways do not play a role, whereas three or more orders of magnitude can be observed at sites characterized by preferential pathways.
A computational fluid dynamics modeling sensitivity analysis reveals that it is not only the presence of contaminant vapor in a preferential pathway that is required to see large observed variations, but there must also exist a permeable region beneath the structure.
Large temporal fluctuations in indoor air contaminant concentrations may be observed where no preferential pathways exist, but this requires a particular combination of large fluctuations in pressure driving force and high soil permeability.
\end{abstract}

@@

Long term vapor intrusion (VI) studies in both residential and larger commercial structures have raised concerns regarding significant observed transient behavior in indoor air contaminant concentrations\cite{u.s._environmental_protection_agency_oswer_2015,folkes_observed_2009,holton_temporal_2013,johnston_spatiotemporal_2014,hosangadi_high-frequency_2017,mchugh_recent_2017,u.s._environmental_protection_agency_assessment_2015}.
VI involves the migration of volatilizing contaminants from soil, groundwater or other subsurface sources into overlying structures. VI has been a recognized problem for some time, but many aspects remain poorly understood, particularly with respect to the causes of large temporal transients in indoor air concentrations.
There is uncertainty within the VI community regarding how to best develop sampling strategies to address this problem\cite{u.s._environmental_protection_agency_oswer_2015,holton_temporal_2013,johnson_integrated_2016}. \par

@@

Similarly, a sewer pipe has recently been suggested to be a source of the contaminants found in the EPA Indianapolis house.
That site was also characterized by large indoor air contaminant concentration fluctuations\cite{mchugh_evidence_2017,u.s._environmental_protection_agency_assessment_2015}.
Sewer lines have been generally implicated as VI sources at several sites\cite{pennell_sewer_2013,mchugh_evidence_2017,roghani_occurrence_2018,riis_vapor_2010}.
A Danish study estimate that roughly 20\% of all VI sites in central Denmark involve significant sewer VI pathways\cite{nielsen_remediation_2017}.
Thus while the consideration of a role of possible sewer or other preferential pathways is now part of normal good practice in VI site investigation\cite{u.s._environmental_protection_agency_oswer_2015}, it is still not known whether the existence of such pathways automatically means that large temporal fluctuations are necessarily to be expected.
In some of these cited cases\cite{pennell_sewer_2013,riis_vapor_2010}, a sewer provided a pathway for direct entry of contaminant into the living space.
While potentially important in many cases, this scenario is not further considered here, where the focus is on pathways that deliver contaminant via the soil beneath a structure. \par

It is, however, now known that even absent a preferential pathway, there may be significant transient variation in indoor air contaminant concentrations at VI sites\cite{folkes_observed_2009,brenner_results_2010,johnston_spatiotemporal_2014}.
One example is a site at NAS North Island at which no preferential pathways have been identified.
Instead, a building at this site is characterized by significant temporal variations in indoor-outdoor pressure differential\cite{hosangadi_high-frequency_2017}.
It is believed that this is the origin of the observed indoor air contaminant concentration fluctuations at that site. \par

This paper investigates the sources of the temporal variation in indoor air contaminant concentrations in both the presence and absence of preferential pathways.
In this work, the latter scenarios are referred to as ”normal” VI scenarios, in which there is typically a groundwater source of the contaminant.
Specifically, we pose the question of just how much variation in indoor air contaminant concentration may be expected at  such normal  VI  sites vs. those characterized by preferential pathways.
The conditions required for preferential pathways to become significant contributors to temporal variations in indoor air contaminant concentrations are also explored, and the consequences for sampling strategies are also discussed.

@@

The ASU house data were obtained over a period of a few years.
During part of this time, controlled pressure method (CPM) tests were being conducted, in which the house was underpressurized to an extent greater than that characterizing “normal” operation.
This caused greater than normal advective flow from the subsurface into the house, thus increasing VI potential\cite{mchugh_evaluation_2012,mchugh_recent_2017,holton_evaluation_2015}.
This period of CPM testing is considered separately from the otherwise ”natural” VI conditions in the analysis.
Likewise, the existence of a preferential pathway at the ASU house needs to be considered in examining the dataset, noting that during some of the testing, this pathway was deliberately cut off, resulting in what we have termed “normal” VI conditions in which the main source of contaminant was believed to be groundwater. \par

@@

Likewise, the Indianapolis site investigation spanned a number of years and periodically included the testing of a sub-slab depressurziation system (SSD).
The goal of the SSD testing was to mitigate the VI risk by drastically depressurizing the sub-slab area underneath the house, preventing the contaminants from entering the structure above.
Only the period before the installation of this system was considered in the present analysis.
It is likely a sewer line beneath the structure acted as a preferential pathway\cite{mchugh_evidence_2017}, however at no point was this preferential pathway removed, making it difficult to assess how significant the role of the preferential pathway was at this site.
Regardless of this it is of interest to consider the data from this site due to how extensive and complete the data collection was. \par

The typical variation in indoor air contaminant concentrations with time will first be considered below in the case of the ASU house during ”natural”, (i.e. non-CPM conditions), in the case of the NAS North Island site over the entire available dataset, and for the Indianapolis case we consider the variations before the installation of the SSD system.
The deviations in indoor air concentration from the mean TCE (and Chloroform and PCE at the Indianapolis site) values, as well as the indoor-outdoor pressure differentials associated with these concentrations were examined.
Both univariate and bivariate kernel density estimations (KDE) were constructed.
KDE is a technique that estimates the probability distribution of a random variable(s) by using multiple kernels, or weighting functions, and in this case, Gaussian kernels are used to create the KDEs.
This means that it is presumed that the variables of interest (i.e., indoor air contaminant concentrations and indoor-outdoor pressure differentials, as sampled) are normally distributed around mean values (and that there are statistical fluctuations associated with each sampling event).
In this instance, the scipy statistical package was used to construct the KDEs, assuming a bandwidth parameter determined by Scott's rule.
The distributions of the individual parameters and the relationship between them will be examined using the KDE method.

\subsection{Modeling Work}
In addition to examining the actual field data, a previously described three -dimensional computational fluid dynamics model of a generic VI impacted house was used to elucidate certain aspects of  the processes.
This model was implemented in a finite element solver package, COMSOL Multiphysics.
In the present work, there has been an addition of a preferential pathway to the ”standard” model that has been described before in publications by this group\cite{shen_influence_2013,yao_investigating_2017,yao_three-dimensional_2017}.
As in the earlier studies, only the vadose zone soil domain is directly modeled.

The modeled structure is assumed to have a 10x10 m foundation footprint, with the bottom of the foundation slab lying 1 m below ground surface (bgs), simulating a house with a basement.
The indoor air space is modeled as a continuously stirred tank (CST)\cite{u.s._environmental_protection_agency_oswer_2015} and all of the contaminant entering the house is assumed to enter with soil gas through a 1 cm wide crack located between the foundation walls and the foundation slab around the perimeter of the house.
All of the contaminant leaving the indoor air space is assumed to do so via air exchange with the ambient.
The indoor control volume is assumed to consist of only of the basement, assumed as having a total volume of $300 \; \mathrm{m^3}$.
Clearly different assumptions could be made regarding the structural features and the size of the crack entry route, but for present purposes, this is unimportant as the intent is only to show for “typical” values what the influence of certain other features can be. \par

The modeled surrounding soil domain extends 5 meters from the perimeter of the house, and is assumed to consist of sandy clay (except as noted).
Directly beneath the foundation slab, there is assumed to be a 30 cm (one foot) thick gravel layer, except in certain cases where this sub-base material is assumed to be the same as the surrounding soil (termed  a ”uniform” soil scenario). \par

Where relevant The preferential pathway is modeled as a 10 cm (4”) pipe that exits into the gravel sub-base beneath the structure.
The air in the pipe is assumed to be contaminated with TCE at a vapor concentration equal to the vapor in equilibrium with the groundwater contaminant concentration below the structure, modified by a scaling factor $\chi$, allowing the contaminant concentration in the pipe to be parameterized. \par

The groundwater beneath the structure is assumed to be homogeneously contaminated with trichloroethylene (TCE) as a prototypical contaminant.
The groundwater itself is not modeled, as the bottom of the model domain is defined by the top of the water table.
The ground surface and the pipe are both assumed to be sources of air to the soil domain.
Both are assumed to be at reference atmospheric pressure. \par

Vapor transport in the soil is governed by Richard’s equation, a modified version  of Darcy’s Law, taking the variability of soil moisture in the vadose zone into account\cite{richards_capillary_1931}.
The van Genuchten equations are used to predict the soil moisture content and thus the effective permeability of the soil\cite{van_genuchten_closed-form_1980}.
The effective diffusivity of contaminant in soil is calculated using the Millington-Quirk model\cite{millington_permeability_1961}.
The transport of vapor contaminant in the soil is assumed to be governed by the advection-diffusion equation, in which either advection or diffusion may dominate depending upon position and particular circumstances.
The equations and the boundary conditions are given in Table \ref{tbl:eqns-bc-parameters}.

@@

\subsection{Drivers For Indoor Air Contaminant Variability}
\subsubsection{The Role Of Building Depressurization/Pressurization}
% KDE figures
\begin{figure}[htb!]
  \caption{KDE analysis of IACC dependence on indoor/outdoor pressure difference at the ASU house and North Island site (\ref{fig:kde-asu-nas}) and the Indianapolis site (\ref{fig:kde-indianapolis}). p-values and Pearson's r-values shown for each dataset.}
  \label{fig:kde-analysis}
  % ASU/North Island plot
  \begin{subfigure}{\textwidth}
    \centering
    \caption{Period before and after the preferential pathway was shut at the ASU house considered separately; North Island dataset considered in its entirety.}
    \label{fig:kde-asu-nas}
    \includegraphics[height=0.4\textheight,keepaspectratio]{nas-asu-house-iacc-deviating.pdf}
  \end{subfigure}
  % Indianapolis plot
  \begin{subfigure}{\textwidth}
    \centering
    \caption{Chloroform and PCE considered separeately at Indianapolis duplex 422.}
    \label{fig:kde-indianapolis}
    \includegraphics[height=0.4\textheight,keepaspectratio]{indianapolis-422.pdf}
  \end{subfigure}
\end{figure}
% Introduction
Of the factors that influence indoor air contaminant concentration IACC in VI, pressure is one of the most dynamic ones, and the relationship between changes in indoor/outdoor pressure difference and IACC are examined in Figure \ref{fig:kde-analysis}.
Three well-studied VI sites are considered; the "ASU house" near Hill AFB in Utah, a building at North Island NAS in California and a duplex in Indianapolis.\par
% Univariate IACC description
The absolute IACCs at these sites vary significantly, therefore some means of comparing them to each other is necessary.
Additionally, the focus in this section is identifying the driver for variations in IACC, and consistent representation of the variations between the different sites must also be achieved.
Here variations on the scale of orders of magnitude is of interest, as these are of greatest concern in proper site IACC characterization.
To achieve these goals, log-10 deviation from the mean IACC (represented as $\mu$) within each dataset is calculated, e.g. $10\mu$ indicates an IACC that is an order of magnitude above the mean IACC for that dataset.
On the y-margins in Figures \ref{fig:kde-asu-nas} and \ref{fig:kde-indianapolis}, the univariate KDE distribution of deviation from the mean may be seen. \par

What these curves on the y-margins characterize are the distributions of sampled IACC at each site.
Naturally, the distributions are centered around the mean ($\mu = 1$).
In same caes, they look close to simple Gaussians in form (e.g. PCE, Tetrachloroethene at Indianapolis).
In most of the other cases, the distributions are curved and not of a simple nature, strongly suggesting that there might be several factors determinig the IACC. \par

% Univariate pressure description
The indoor/outdoor pressure differences are simply shown as reported, and their univariate KDE distributions are shown on the x-margins.
A negative pressure difference indicates that the building is underpressurized relative to ambient.\par
% Bivariate description
The relationship between the deviation from mean IACC, and indoor/outdoor pressure difference may be seen in the central portion of each figure, in the form of a bivariate KDE distribution.
The p-value and Pearson's r-value for each bivariate distribution are shown in the legend.\par
% Sub-figure description (details about sites)
In Figure \ref{fig:kde-asu-nas} the North Island and ASU house datasets are plotted, with blue representing the North Island site, and with red and green representing the ASU house.
The ASU house dataset is split into two different parts, the first (red) are data from the time period before the preferential pathway was discovered, the second (green) is from after the preferential pathway was sealed off.
At both the ASU house and North Island sites, TCE was the only contaminant considered.
The IACC of Chloroform and PCE in the 422 part of the Indianapolis duplex were considered in Figure \ref{fig:kde-indianapolis}.\par
% North Island discussion
From Figure \ref{fig:kde-asu-nas}, it is apparent that the three datasets differ significantly from each other.
The North Island site exhibit the greatest variation in both pressure difference and IACC, in fact, looking at the univariate pressure distribution - it is almost entirely flat and spanning from extremly high and low values.
A likely explation for this is the poor condition of the structure, rendering it highly susceptible to envivornmental influnces.\par

To accompany this, there is significant variation in IACC, that has a relativly strong link with pressure difference ($r = -0.71$), although there is clearly a distinct peak (around $0.5\mu$) that is primally associated with the smaller negative and positive pressure differences ($~-4 \leq \Delta p \leq 8$).
This demonstrates that at this particular site, the very large variations (an order of magnitude or more) are primarily influenced by the unusually large pressure difference fluctations.
Likely, this relationship is amplified by the fact that the soil surrounding the structure is sandy and therefore relatively permeable - increasing the influence of pressure differences.
However, despite all this, pressure differences can only partly explain the variations observed at this site.\par
% ASU discussion
The two ASU datasets are not only significantly different from the North Island dataset, but also different from each other.
Consider the dataset before the preferential pathway was discovered (red) and one can see that the IACC varies significantly (from $0.05 \mu$ to more than $50 \mu$, albeit rarely.)
This is more variation than was observed at North Island, and yet the distribution of pressure differences is narrow, mostly varying by a few Pa around zero (~$\pm 2$ Pa), and with $r = -0.50$, suggests that pressure difference is not insignificant but still a weak predictor for the variations in IACC.\par

The picture changes completely after the preferential pathway was closed off, where a similar distribution of pressure differences is observed, but with a signficant reduction in IACC variance to around $\pm 0.5 \mu$, resembling something of a log-normal distribution.
This clearly shows just how much variance the preferential pathway at this site contributed with, and how critical it is to assess whether a preferential pathway is present at a particular site in general, lest potentially face unacceptable uncertainty in determining the relevant contaminant exposure.
It is also clear that in the absence of a preferential pathway, the pressure difference becomes even less important, with the p- and r-value indicating that pressure differences are here insignficant in determining variation in IACC.
This also suggests that the preferential pathway at the ASU house acted not only as a preferential pathway for contaminant vapors, but for air in general - allowing the impermeable soil surrounding the house to be circumvented.
preferential pathways, and the factors that influnce them will be explored later in this article.\par
% Indianapolis discussion
The Indianapolis duplex also featured a preferential pathway, but despite this, there is significantly less variance in IACC of both Chloroform and PCE.
The exact reason for this is not known, but considering that the preferential pathway at the Indianapolis site was a sanitary sewer, and at ASU the preferential pathway was a foundation drain system, it is not inconceivable that their dynamics are quite different.
Nevertheless, both PCE and Chloroform are more or less log-normally distributed, with (mostly) less than an order of magnitude variance around the mean, indicating that the varation in IACC are species independent.\par

The pressure difference distribution exhibit some bimodal characteristics, with a larger variation than was observed at the ASU site.
This may explain why the r-values for both PCE and Chloroform, $r=-0.51$ and $r=-0.53$ respectively, are so similar to the ASU site before the preferential pathway there was closed; the larger pressure differences, which occured more often than ASU, increase contaminant entry rates to the extent that a preferential pathway was needed to "achieve" at the ASU site.\par
% Summary discussion
All of these data clearly suggest that there is significant variation in IACC that cannot be easily explained by variations in indoor/outdoor pressure difference - even with a preferential pathway.
It is also clear that preferential pathway may contribute with unacceptable levels of variations, and screening for preferential pathways should be part of any site investiation.
It may also be prudent, even in the absence of preferential pathway, to add around a half to one order of magnitude margin of error to early screening measurements of indoor air.
E.g. if a single sample is less than an order of magnitude from some target limit, it may be justified to perform subsequent samples to better assess the relevant exposure.
These data also call the role of advection in VI into question, as "normal" pressure difference ranges are such a poor predictor of IACC variation, which is the opposite of what one would expect if advection was a dominant transport mechanism.\par

@@

\begin{figure}[htb!]
  \caption{Sensitivity analysis of IACC dependence on indoor/outdoor pressure difference for cases featuring a preferential pathway (\ref{fig:ss-sensitivity-analysis-pp}) and without a preferential pathway (\ref{fig:ss-sensitivity-analysis-no-pp}). Results compared with field data from ASU house.}
  \label{fig:ss-sensitivity-analysis}
  % "Overview" plot
  \begin{subfigure}{0.55\textwidth}
    \caption{preferential pathway present. Sensitivity to the presence of a gravel sub-base and contamination in the preferential pathway considered.}
    \label{fig:ss-sensitivity-analysis-pp}
    \includegraphics[width=\textwidth]{yes-pp.pdf}
  \end{subfigure}
  % Violinplot figure
  \begin{subfigure}{0.55\textwidth}
    \caption{No preferential pathway present. Sensitivitiy to the presence of a gravel sub-base considered.}
    \label{fig:ss-sensitivity-analysis-no-pp}
    \includegraphics[width=\textwidth]{no-pp.pdf}
  \end{subfigure}
\end{figure}

Clearly, preferential pathways pose a major problem for site investigations, in particular because they may be difficult or even impossible to anticipate or uncover; even at a well studied site like the ASU house it took years to discover the preferential pathway.
Therefore, there is a need to consider other indications that a preferential pathway is present or a potential issue.\par

To investigate this, inspiration is taken from the ASU house, where there clearly was a preferential pathway containing contaminant vapor, which exited into a gravel (permeable) sub-base.
The impact of these features is simulated by examining the IACC at different indoor/outdoor pressure differences for each combination of cases.
Instead of representing IACC as absolute concentration, we now non-dimensionalize with respect to the vapor in equilibrium with the groundwater contaminant concentration.
This non-dimensionalized property is commonly called the attenuation factor and is denoted by $\alpha_\mathrm{gw}$.
The result of a steady-state sensitivity analysis relative to several factors is seen in Figure \ref{fig:ss-sensitivity-analysis}.\par
